---
title: "Level orders of a binary tree"
description: "A solution (kind of) to leetcode problem 102"
author:
  - name: Rashmin Chitale
    url: 
date: 04-22-2025
categories: [leetcode,R] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(R6)
```

# The background

::: callout-note
## When are level orders of a binary tree?

The level orders of a tree are the order of each node at each level (depth) of a binary tree (going in a left to right fashion).
:::

It is easy to understand what this is with an example.

```{mermaid}
%%| fig-align: center
graph TD
  A((01)) --> B((02))
  A((01)) --> C((03))
  B((02)) --> D((04))
  B((02)) --> E((05))
  C((03)) --> F((06))
  C((03)) --> G((07))
  
```

For this above tree we would output the following `[1], [2,3], [4,5,6,7]`

```{mermaid}
%%| fig-align: center
graph TD
  A((01)) --> B((02))
  A((01)) --- C((03)):::hidden
  B((02)) --> D((04))
  B((02)) --> E((05))
  E((05)) --> F((06))
  E((05)) --> G((07))

  %% Make link B-->C invisible (this is the second link, index 1)
  linkStyle 1 stroke:transparent,stroke-width:0
  
```

And for this tree we would output `[1], [2], [4,5], [6,7]`

# The solution

In my solution, I will carry over some of the code from my other [post](../2025-01-16-leetcode-max-depth-tree/index.qmd). I first define a node as follows

```{r}
Node = R6Class(
  classname = "Node", 
  public = list(
    left = NA,
    right = NA,
    value = "numeric",
    initialize =
      function(left = NA, right = NA, value = NA){
        self$left = left
        self$right = right
        self$value = value
      }
  )
)

```

The main algorithm is then as follows

```{r}
LTree = R6Class(
  classname = "Tree",
  public = list(
    depth = "numeric",
    dat = "data.frame",
    initialize = 
      function(depth, dat){
        self$depth = -Inf
        self$dat = data.frame(depth = c(NA), value = c(NA), index = c(NA))
      },
    node.df =
      function(tree, index = 1, cdepth = 0){
        
        #increase current depth by 1
        up.depth = cdepth + 1
        
        #compare to global depth and increase if larger
        if(up.depth > self$depth){
          self$depth = up.depth
        }
        
        #add root node at depth 1 to df
        if(up.depth == 1){
          self$dat[1, ] = c(1,tree[["value"]],1)
        }
  
        #check for nullity of children nodes
        left.na = is.environment(tree[["right"]]) & !is.environment(tree[["left"]])
        right.na = !is.environment(tree[["right"]]) & is.environment(tree[["left"]])
        both.na = !is.environment(tree[["right"]]) & !is.environment(tree[["left"]])
        
        #array index of children node
        left.index = 2*index
        right.index = 2*index + 1
        
        #choose what type of recursion to carry out
        if(left.na == TRUE){
          self$dat[nrow(self$dat)+1,] = c(up.depth+1,tree[["right"]][["value"]],right.index)
          self$node.df(tree$right, cdepth = up.depth, index = right.index)
        } else if(right.na == TRUE){
          self$dat[nrow(self$dat)+1,] = c(up.depth+1,tree[["left"]][["value"]],left.index)
          self$node.df(tree$left, cdepth = up.depth, index = left.index)
        } else if (both.na == TRUE){
          return()
        } else {
          self$dat[nrow(self$dat)+1,] = c(up.depth+1,tree[["right"]][["value"]],right.index)
          self$dat[nrow(self$dat)+1,] = c(up.depth+1,tree[["left"]][["value"]],left.index)
          self$node.df(tree$right, cdepth = up.depth, index = right.index)
          self$node.df(tree$left, cdepth = up.depth, index = left.index)
        }

      },
    sort.ind =
      function(value, index){
        #this function sorts a vec of values, by the index values in another vec
        ord.vals = double(length = length(value))
        for(i in 1:length(index)){
          min.ind = which(index  == min(index))
          ord.vals[[i]] = value[min.ind]
          index = index[-min.ind]
          value = value[-min.ind]
          next(i)
        }
        return(ord.vals)
      },
    level.traverse =
      function(tree){
        #this function takes the df containing node information, and sorts it
        self$node.df(tree)
        tree.levels = list()
        for(i in 1:self$depth){
        dat.subset = self$dat[which(self$dat$depth == i), ]
        tree.levels[[i]] = self$sort.ind(dat.subset$value, dat.subset$index)
        next(i)
        }
        return(tree.levels)
      }
  
  ))
```

The algorithm works as expected on both of the trees we used as an example earlier.

```{r}
tree1 = Node$new(value = 1)
tree1[["left"]] = Node$new(value = 2)
tree1[["right"]] = Node$new(value = 3)
tree1[["left"]][["left"]] = Node$new(value = 4)
tree1[["left"]][["right"]] = Node$new(value = 5)
tree1[["right"]][["left"]] = Node$new(value = 6)
tree1[["right"]][["right"]] = Node$new(value = 7)

LTree$new()$level.traverse(tree1)
```

```{r}
tree2 = Node$new(value = 1)
tree2[["left"]] = Node$new(value = 2)
tree2[["left"]][["left"]] = Node$new(value = 4)
tree2[["left"]][["right"]] = Node$new(value = 5)
tree2[["left"]][["right"]][["left"]] = Node$new(value = 6)
tree2[["left"]][["right"]][["right"]] = Node$new(value = 7)

LTree$new()$level.traverse(tree2)


```




