---
title: "Danny's Diner"
description: "My solutions to the Danny's Diner case study"
author:
  - name: Rashmin Chitale
    url: 
date: 04-05-2026
categories: [sql]
---

# Introduction

Danny's Diner is a case study made available as part of the 8 week SQL challenge by Danny Ma. It includes a set of related tables on the transactions of customers at a diner. You can find this case study at this [link](https://8weeksqlchallenge.com/). I will be using the case study to revise SQL, which I have not used in a while.

# Setup

```{r warnings = F, error=F, message = F, echo = T, output = F}

# Required packages

library(duckdb)
library(DBI)
library(tibble)
library(lubridate)

#Create a database/connection in memory

con = DBI::dbConnect(duckdb::duckdb(), dbdir = ':memory:')

#Create tables in R

sales = tibble(
    customer_id = c('A','A','A','A','A','A','B','B',
                    'B','B','B','B','C','C','C'),
    order_date = mdy('1/1/2021','1/1/2021','1/7/2021','1/10/2021',
                     '1/11/2021','1/11/2021','1/1/2021','1/2/2021',
                     '1/4/2021','1/11/2021','1/16/2021','2/1/2021',
                     '1/1/2021','1/1/2021','1/7/2021'),
    product_id = c(1,2,2,3,3,3,2,2,1,1,3,3,3,3,3))

menu = tibble(
  product_id = c(1,2,3), 
  product_name = c('sushi','curry','ramen'), 
  price = c(10,15,12))

members = tibble(
  customer_id = c('A','B'),
  join_date = mdy('1/7/2021', '1/9/2021'))

#Register connection of tables in R to the SQL database in memory

duckdb_register(con, "sales", sales)
duckdb_register(con, "menu", menu)
duckdb_register(con, "members", members)

```

# Questions

## Question 1

```{sql}
#| connection: con

/* Total amount spent by each customer */ 

SELECT customer_id, SUM(price) AS total_spent
FROM sales
INNER JOIN menu ON menu.product_id = sales.product_id
GROUP BY customer_id
```

```{r eval = F, include = F}
dbGetQuery(con, 'SELECT customer_id, COUNT(DISTINCT order_date) AS days_visited FROM sales GROUP BY customer_id')
```

## Question 2

```{sql}
#| connection: con

/* Days visited by each customer */ 

SELECT customer_id, COUNT(DISTINCT order_date) AS days_visited
FROM sales
GROUP BY customer_id
```

## Question 3

```{sql}
#| connection: con

/* First item from menu purchased by each customer */ 

SELECT product_name, order_date, customer_id
FROM sales
INNER JOIN menu ON sales.product_id = menu.product_id
WHERE order_date IN (
  SELECT MIN(order_date)
  FROM sales
  GROUP BY customer_id)

```

## Question 4

```{sql}
#| connection: con

/* Most purchased item on menu, and no of times it was purchase */

SELECT product_name, COUNT(customer_id) as total_purchased
FROM sales
INNER JOIN menu ON sales.product_id = menu.product_id
GROUP BY product_name
```

We want to filter the above table using sub-queries

```{sql}
#| connection: con

/* Most purchased item on menu, and no of times it was purchase */

SELECT product_name, COUNT(customer_id) as total_purchased
FROM sales
INNER JOIN menu ON sales.product_id = menu.product_id
GROUP BY product_name
HAVING /* Filter the original table for item which had max purchases */ 
  total_purchased 
  IN 
  ( /* This sub-query gets number of times most purchased item was bought */ 
  SELECT MAX(total_purchased)
  FROM 
  ( /* This sub-query gets product and number of purchases */ 
  SELECT product_name, COUNT(customer_id) as total_purchased
  FROM sales
  INNER JOIN menu ON sales.product_id = menu.product_id
  GROUP BY product_name) AS purchases
  )

```

## Question 5

```{sql}
#| connection: con

/* Most popular item for each customer */

SELECT DISTINCT customer_id, product_name,
  COUNT() OVER( PARTITION BY product_name, customer_id ) AS times_bought
FROM sales 
INNER JOIN menu ON sales.product_id = menu.product_id
ORDER BY sales.customer_id, menu.product_name ASC

```

From this table we want to filter for the most popular item for each customer (same popularity if times bought is tied)

```{sql}
#| connection: con

/* Most popular item for each customer */

SELECT t1.customer_id, t2.product_name, t2.times_bought
FROM(
  /* On the table of preferences we find number of times they bought their most fav item*/ 
  SELECT DISTINCT customer_id, 
         MAX(times_bought) OVER (PARTITION BY customer_id) AS max_bought
  FROM(
    /* Table of preferences for each customer */ 
    SELECT DISTINCT customer_id, product_name,
    COUNT() OVER( PARTITION BY product_name, customer_id ) AS times_bought
    FROM sales 
    INNER JOIN menu ON sales.product_id = menu.product_id
    )) AS t1
/* We now inner join t1 and t2 tables based off a 2 column key */ 
INNER JOIN (
  /* This is the same as the table of preference from above */ 
  SELECT DISTINCT customer_id, product_name,
  COUNT() OVER( PARTITION BY product_name, customer_id ) AS times_bought
  FROM sales 
  INNER JOIN menu ON sales.product_id = menu.product_id
  ) AS t2 
  ON t1.customer_id = t2.customer_id AND 
     t1.max_bought = t2.times_bought 
ORDER BY t1.customer_id ASC


```

## Question 6

```{sql}
#| connection: con

/* Item first bought after becoming a member */

SELECT t1.customer_id, t2.product_name, t2.order_date
FROM (
  SELECT /* Subquery takes result of below subquery and finds the first purchase after becoming a member*/ 
    DISTINCT customer_id, MIN(order_date) OVER(PARTITION BY customer_id) AS after_member
  FROM(
    /* Below subquery gets all purchases for each member >= join date */ 
    SELECT sales.customer_id, menu.product_name, order_date
    FROM sales
    INNER JOIN members ON
      sales.customer_id = members.customer_id AND
      sales.order_date >= members.join_date
    INNER JOIN menu ON
      sales.product_id = menu.product_id)
) AS t1
INNER JOIN ( /* Match customer id and first purchase in the 2 tables */
  SELECT sales.customer_id, menu.product_name, order_date
  FROM sales
  INNER JOIN members ON
    sales.customer_id = members.customer_id AND
    sales.order_date >= members.join_date
  INNER JOIN menu ON
    sales.product_id = menu.product_id
) AS t2
ON t1.customer_id = t2.customer_id AND
   t1.after_member = t2.order_date

```

## Question 7

```{sql}
#| connection: con

/* Item bought just before becoming a member */

SELECT t1.customer_id, t2.product_name, t2.order_date
FROM (
  SELECT /* Subquery takes result of below subquery and finds the first purchase after becoming a member*/ 
    DISTINCT customer_id, MAX(order_date) OVER(PARTITION BY customer_id) AS after_member
  FROM(
    /* Below subquery gets all purchases for each member >= join date */ 
    SELECT sales.customer_id, menu.product_name, order_date
    FROM sales
    INNER JOIN members ON
      sales.customer_id = members.customer_id AND
      sales.order_date < members.join_date
    INNER JOIN menu ON
      sales.product_id = menu.product_id)
) AS t1
INNER JOIN ( /* Match customer id and first purchase in the 2 tables */
  SELECT sales.customer_id, menu.product_name, order_date
  FROM sales
  INNER JOIN members ON
    sales.customer_id = members.customer_id AND
    sales.order_date < members.join_date
  INNER JOIN menu ON
    sales.product_id = menu.product_id
) AS t2
ON t1.customer_id = t2.customer_id AND
   t1.after_member = t2.order_date

```

## Question 8

```{sql}
#| connection: con

/* Number of purchases and total spent before becoming member */

SELECT sales.customer_id, SUM(price) AS pre_member_spent, COUNT() AS pre_member_purch_num
FROM sales
INNER JOIN members ON
  sales.customer_id = members.customer_id AND
  sales.order_date < members.join_date
INNER JOIN menu ON
  sales.product_id = menu.product_id
GROUP BY sales.customer_id

```

## Question 9

```{sql}
#| connection: con
#| max.print: -1

/* Points earned; 10 for each dollar, times 2 for sushi */ 

SELECT customer_id, SUM(points) AS total_points
FROM(
  /* Create multiplier, points for each product */ 
  SELECT sales.customer_id,
  CASE product_name
    WHEN 'sushi' THEN 20
    ELSE 10
  END AS multiplier, price*multiplier AS points
  FROM sales
  INNER JOIN menu ON
    sales.product_id = menu.product_id)
GROUP BY customer_id

```

## Question 10

```{sql}
#| connection: con
#| max.print: -1

/* Points earned; times 2 for 1 week after becoming member */ 

/* Subquery in a CTE to create points earned on each order based off conditionals */
WITH trans_points_cte AS(
    SELECT sales.customer_id, 
         sales.order_date, members.join_date,
         menu.product_name, menu.price,
         /* End date of membership points benefit */
         join_date + 7 AS end_mult_date,
         /* Logical when benefit active based off relative order date */ 
         order_date BETWEEN join_date AND end_mult_date AS in_mult,
         /* Multiplier for each dollar amount */ 
         CASE
           WHEN in_mult = TRUE THEN 20 
           WHEN product_name = 'sushi' THEN 20
           ELSE 10
         END AS multiplier,
         /* Points is price times multiplier */
         price*multiplier AS points
  FROM sales
  /* Inner join all 3 tables */ 
  INNER JOIN members ON
    sales.customer_id = members.customer_id
  INNER JOIN menu ON
    sales.product_id = menu.product_id
  /* Filter for orders in january only */ 
  WHERE order_date < '2021-02-01'
)

/* Sum points for each customer */
SELECT customer_id, SUM(points) AS total_points
FROM trans_points_cte
GROUP BY customer_id 
```

## Additional Question 1

```{sql}
#| connection: con
#| max.print: -1

/* Table with additional column for member */ 

SELECT sales.customer_id, sales.order_date,
       menu.product_name, menu.price,
       CASE
         WHEN join_date IS NULL THEN 'N'
         ELSE 'Y'
       END AS member
FROM sales
LEFT OUTER JOIN members ON
  sales.customer_id = members.customer_id
INNER JOIN menu ON
  sales.product_id = menu.product_id
```

## Additional Question 2

```{sql}
#| connection: con
#| max.print: -1

/* Rankings (date) of purchases made after becoming a member */

WITH whole_table_cte AS(
  SELECT sales.customer_id, order_date,
         product_name, price,
         join_date,
         order_date >= join_date AS member,
         CASE
           WHEN member = TRUE THEN 
             DENSE_RANK() OVER(PARTITION BY sales.customer_ID, member 
                               ORDER BY order_date ASC)
           ELSE 'null'
         END AS ranking
  FROM sales
  INNER JOIN menu ON
    sales.product_id = menu.product_id
  LEFT OUTER JOIN members ON
    sales.customer_id = members.customer_id
  ORDER BY sales.customer_id ASC
)

SELECT customer_id, order_date, product_name, price, member, ranking
FROM whole_table_cte
```
